---
title: 'P8105: Homework 2'
author: "Derek Lamb"
date: "`r Sys.Date()`"
output: github_document
---
Load all relevant packages.
```{r load packages}
library(tidyverse)
```

# Problem 1
I loaded in the dataframe and separated out the mon variable as indicated. Because the prez_gop and prez_dem data were redundant, I based the new president variable on prez_gop, then used `select` to remove unneeded variables.
```{r import data p1}
# Load in the data set and separate the mon variable into year, month, day

df_pol <- read_csv("data/pols-month.csv") |> 
  separate(mon, into = c("year", "month", "day"), "-") |> 
  mutate(
    month = month.name[as.numeric(month)],
    president = case_match(
      prez_gop,
      1 ~ "gop",
      0 ~ "dem"
    )) |> 
  select(-c(day, prez_dem, prez_gop))



```

I read in `snp.csv` and went through the same cleaning steps. The date was formatted differently, so I used `lubridate` to reformat it. However, this assigned the incorrect year to everything before 1969, so I used `case_when` to reassign the year.
```{r}
df_snp <- read_csv("data/snp.csv") |> 
  mutate(date = lubridate::mdy(date)) |> 
  separate(date, into = c("year", "month", "day"), "-") |> 
  mutate(
    month = month.name[(as.numeric(month))],
    year = as.numeric(year),
    year = case_when(
      year > 2020 ~ year - 100,
      year < 2020 ~ year
    )
    ) |> 
  select(-day) |> 
  arrange(year)

df_snp
```

