---
title: 'P8105: Homework 2'
author: "Derek Lamb"
date: "`r Sys.Date()`"
output: github_document
---
Load all relevant packages.
```{r load packages, message = FALSE}
library(tidyverse)
library(readxl)
library(lubridate)
```

# Problem 1
I loaded in the dataframe and separated out the mon variable as indicated. Because the prez_gop and prez_dem data were redundant, I based the new president variable on prez_gop, then used `select` to remove unneeded variables.
```{r import pols data, message = FALSE}
# Load in the data set and separate the mon variable into year, month, day

df_pol <- read_csv("data/pols-month.csv") |> 
  separate(mon, into = c("year", "month", "day"), "-") |> 
  mutate(
    month = month.name[as.numeric(month)],
    year = as.numeric(year),
    president = case_match(
      prez_gop,
      1 ~ "gop",
      0 ~ "dem"
    )) |> 
  select(-c(day, prez_dem, prez_gop))
```

I read in `snp.csv` and went through the same cleaning steps. The date was formatted differently, so I used `lubridate` to reformat it. I ended up using `lubridate` more later, so I added it to the code chunk to load packages. However, this assigned the incorrect year to everything before 1969, so I used `case_when` to reassign the year.
```{r import snp data, message = FALSE}
df_snp <- read_csv("data/snp.csv") |> 
  mutate(date = mdy(date)) |> 
  separate(date, into = c("year", "month", "day"), "-") |> 
  mutate(
    month = month.name[(as.numeric(month))],
    year = as.numeric(year),
    year = case_when(
      year > 2020 ~ year - 100,
      year < 2020 ~ year)
    ) |> 
  rename(snp_close = close) |> 
  select(-day) |> 
  arrange(year)

```

In working with the unemployment data, I decided to convert the abbreviated months to full months by mapping from the built in vector `month.abb` to another built in vector `month.name`. I had used `month.name` above to go from month numbers to month names.
```{r import unemployment data, message = FALSE}
df_unemp <- read_csv("data/unemployment.csv") |> 
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "unemployment"
  ) |> 
  mutate(month = month.name[match(month, month.abb)]) |> 
  janitor::clean_names()
```

Now I will merge these three dataframes into one dataframe.

```{r}
# First, left_join political and s&p data
df_combined <- left_join(df_pol, df_snp, by = join_by(year, month))

# Next, add unemployment data
df_combined <- left_join(df_combined, df_unemp, by = join_by(year, month))
```

The `df_combined` dataframe is contains `r nrow(df_combined)` rows of `r ncol(df_combined)` variables. The dataframe contains monthly records of U.S. political control at the state and federal level, particularly whether the `president` was a democrat or republican, the S&P 500 index (`snp_close`), and unemployment rate (`unemployment`). The political data ranges from 1947-2015, although the S&P 500 and unemployment data only go back to 1950 and 1948, respectively, so the first 36 rows of the dataframe contain missing data.

# Problem 2
I already loaded `readxl`, so I will now import the Mr. Trash Wheel data. The first row is an image, so I will skip that. There are two extra columns that get read in, and I remove both of them.
```{r mr. trash wheel, message = FALSE}
# Import and clean the Mr. Trash Wheel data
df_mr_tw <- read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", skip = 1) |> 
  janitor::clean_names() |> 
  rename(weight = weight_tons, volume = volume_cubic_yards) |> 
  select(-x15, -x16) |> 
  filter(dumpster != "") |> 
  mutate(homes_powered = weight*500/30) |> 
  mutate(wheel = "mr_trash_wheel",
         year = as.numeric(year)) |> 

# There is a typo in the 'date' variable of the data frame
# I decided to correct it here

  mutate(date = case_when(
    date == ymd("1900-01-20") ~ ymd("2020-01-20"),
    date != ymd("1900-01-20") ~ date))
```

I then read in the Professor and Gwynnda Trash Wheel data, noting that `df_prof_tw` does not have the variable `sports_balls` and that `df_gwyn_tw` does not have the variables `glass_bottles`, `chip_bags`, or `sports_balls`. While Mr. and Professor Trash Wheels containt the variable `grocery_bags`, Gwynnda contains the variable `plastic_bag`. I renamed this for consistencies of variable names.

```{r other trash wheels}
# Import and clean Prof. Trash Wheel data
df_prof_tw <- read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", skip = 1) |> 
  janitor::clean_names() |> 
  rename(weight = weight_tons, volume = volume_cubic_yards) |> 
  filter(dumpster != "") |> 
  mutate(homes_powered = weight*500/30) |> 
  mutate(wheel = "prof_trash_wheel")

# Import and clean Gwynnda Trash Wheel data
df_gwyn_tw <- read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", skip = 1) |> 
  janitor::clean_names() |> 
  rename(weight = weight_tons, 
         volume = volume_cubic_yards,
         grocery_bags = plastic_bags) |> 
  filter(dumpster != "") |> 
  mutate(homes_powered = weight*500/30) |> 
  mutate(wheel = "gwynnda_trash_wheel")
```

I then read in  that `df_prof_tw` does not have the variable `sports_balls` and that `df_gwyn_tw` does not have the variables `glass_bottles`, `chip_bags`, or `sports_balls`. 

Now I will join the dataframes together.

```{r join dataframes}

df_tw <- bind_rows(df_mr_tw, df_prof_tw) |> 
  bind_rows(df_gwyn_tw) |> 
  select(-dumpster) |> 
  select(year, month, wheel, everything()) |> 
  arrange(date)

```

The `df_tw` dataframe contains `r nrow(df_tw)` observations of `r ncol(df_tw)` variables for `r length(unique(pull(df_tw, wheel)))` different trash wheels. On each `date`, the `weight` in tons and `volume` in cubic yards of trash collected was recorded, as well as the counts of certain common trash items, such as `cigarette_butts` or `plastic_bottles`. This dataframe can be used to answer questions about the performance of the different trash wheels. For example, Professor Trash Wheel has collected a total of `r with(df_tw, sum(weight[wheel == "prof_trash_wheel"]))` tons of trash. Gwynnda the Trash Wheel, which started operating in `r with(df_tw, min(year[wheel == "gwynnda_trash_wheel"]))`, collected `r format(with(df_tw, sum(cigarette_butts[wheel =="gwynnda_trash_wheel" & year == 2021 & month == "July"])), scientific = FALSE, OutDec = ",")` in July of that year.


